Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades

### 2022-11-07 notebooks/expected_and_actual_counts_ratio.Rmd
Compare synonymous and non-synonymous mutations on a regular scale

#install.packages("data.table")
#install.packages("ggplot2")
#install.packages("stringr")
#install.packages("gtools")

library(data.table)
library(ggplot2)
library(stringr)
library(gtools)
```

The purpose of this notebook is to generate volcano plots comparing any two SARS-CoV-2 phylogenetic clades. Using the volcano plots, mutations at certain nucleotide sites that are undergoing very different directions or degrees of selection between the two clades can be identified. Once this notebook has been finalized, sections of the code will be saved as scripts and run through a Snakemake pipeline, allowing the 78 clade comparisons to be computed in parallel.

@@

# Identify the two clades that will be included in comparisons.
clades <- c("21J","22B")

@@

```{r subset_by_clade}
# Create a subset of the data using only the clades that are going to be compared.
filtered_clades <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
```

@@

```{r restructure_data}
# Create two data frames with data for clade 1 and clade 2.
clade1 <- filtered_clades[which(filtered_clades$clade==clades[1]),c("nt_mutation","clade","actual_count","rounded_expected","observed_expected","synonymous","four_fold_degenerate")]
clade2 <- filtered_clades[which(filtered_clades$clade==clades[2]),c("nt_mutation","clade","actual_count","rounded_expected","observed_expected","synonymous","four_fold_degenerate")]

# Append 'clade1' or 'clade2' to the column names so that the two data frames can be combined with easy to interpret columns.
colnames(clade1)[2:ncol(clade1)] <- paste("clade1", colnames(clade1)[2:ncol(clade1)], sep="_")
colnames(clade2)[2:ncol(clade2)] <- paste("clade2", colnames(clade2)[2:ncol(clade2)], sep="_")
colnames(clade1)[which(colnames(clade1)=="clade1_clade")] <- "clade1"
colnames(clade2)[which(colnames(clade2)=="clade2_clade")] <- "clade2"

# Merge the two data frames by mutation so that each row represents one comparison.
comparisons <- merge(clade1, clade2, by="nt_mutation")
```

@@

```{r contingency_table}
# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["clade1_actual_count"]),as.numeric(c["clade1_rounded_expected"])),c(as.numeric(c["clade2_actual_count"]),as.numeric(c["clade2_rounded_expected"]))))$p.value}))

@@

```{r p_value_correction}
# Using Bonferroni Correction, adjust the p-values by the number of hypothesis tests.
comparison_p_values$p_value_corrected <- comparison_p_values$p_value*nrow(comparison_p_values)
```

@@

```{r fold-change}
# Calculate fold change for plotting volcano plots.
comparison_p_values$fold_change <- log2(comparison_p_values$clade1_observed_expected/comparison_p_values$clade2_observed_expected)
```

@@

```{r volcano_plot, echo=FALSE}
ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value_corrected)))+
  geom_point()+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  theme_bw()
```

@@

To verify the analyses are generating expected results, compare the p-values for synonymous mutations to non-synonymous mutations. Unlike non-synonymous mutations, synonymous mutations should not be experiencing selective pressures. Therefore, while there might be differences in the direction and degree of selection on non-synonymous (i.e., p-value close to 0), there should be insignificant differences between clades on synonymous mutations (i.e., p-values close to 1).

```{r verify_analyses}
# sanity check analyses (distribution of synonymous p-values)
verify_analyses <- comparison_p_values
verify_analyses$synonymous <- paste(comparison_p_values$clade1_synonymous, comparison_p_values$clade2_synonymous, sep=".")

ggplot(verify_analyses, aes(x=synonymous, y=-log10(p_value_corrected)))+
  geom_jitter(alpha=0.5)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)
```

>>>>

#install.packages("data.table")
#install.packages("ggplot2")
#install.packages("ggforce")
#install.packages("stringr")
#install.packages("gtools")

library(data.table)
library(ggplot2)
library(ggforce)
library(stringr)
library(gtools)
```

The purpose of this notebook is to generate a volcano plot comparing any two SARS-CoV-2 phylogenetic clades. Using the volcano plot, mutations at certain nucleotide sites that are undergoing very different directions or degrees of selection between the two clades can be identified. Once this notebook has been finalized, sections of the code will be saved as scripts and run through a Snakemake pipeline, allowing all clade comparisons to be computed in parallel.

@@

# Identify the two clades that will be included in comparisons.
clades <- c("20A","22B")

@@

```{r subset_by_clade}
# Create a subset of the data using only the clades that are going to be compared.
filtered_clades <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]

# Remove columns that were already used to filter the data or than contain redundant information.
filtered_clades$subset <- NULL
filtered_clades$exclude <- NULL
filtered_clades$nt_site <- NULL
```

@@

```{r restructure_data}
# Create two data frames with data for clade 1 and clade 2.
clade1 <- filtered_clades[which(filtered_clades$clade==clades[1]),]
clade2 <- filtered_clades[which(filtered_clades$clade==clades[2]),]

# Append 'clade1' or 'clade2' to the column names so that the two data frames can be combined with easy to interpret columns.
colnames(clade1) <- paste("c1", colnames(clade1), sep="_")
colnames(clade2) <- paste("c2", colnames(clade2), sep="_")

# Modify nt_mutation, clade_founder_nt and clade to make merging easier and make the data frame more readable.
{colnames(clade1)[which(colnames(clade1)=="c1_nt_mutation")] <- "nt_mutation"
colnames(clade2)[which(colnames(clade2)=="c2_nt_mutation")] <- "nt_mutation"
colnames(clade1)[which(colnames(clade1)=="c1_clade_founder_nt")] <- "clade_founder_nt"
colnames(clade2)[which(colnames(clade2)=="c2_clade_founder_nt")] <- "clade_founder_nt"
colnames(clade1)[which(colnames(clade1)=="c1_clade")] <- "clade1"
colnames(clade2)[which(colnames(clade2)=="c2_clade")] <- "clade2"}

# Merge the two data frames by mutation so that each row represents one comparison.
comparisons <- merge(clade1, clade2, by=c("nt_mutation","clade_founder_nt"))
```

@@

```{r contingency_table}
# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["c1_actual_count"]),as.numeric(c["c1_rounded_expected"])),c(as.numeric(c["c2_actual_count"]),as.numeric(c["c2_rounded_expected"]))))$p.value}))

@@

```{r p_value_correction}
# Using False Discovery Rate Correction, adjust the p-values by the number of hypothesis tests.
comparison_p_values$p_value_corrected <- p.adjust(comparison_p_values$p_value, method="fdr", n=nrow(comparison_p_values))
```

@@

```{r fold-change}
# Calculate fold change for plotting volcano plots.
comparison_p_values$fold_change <- log2(comparison_p_values$c1_observed_expected/comparison_p_values$c2_observed_expected)
```

@@

```{r volcano_plot, echo=FALSE}
# Plot the uncorrected p-values.
ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value)))+
  geom_point(alpha=0.5)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  theme_bw()

# After dropping p = 1, plot FDR corrected p-values (color by nucleotide clade founder and label by mutation).
volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=nt_mutation, color=clade_founder_nt))+
  geom_point()+
  geom_text(color="black", size=2, nudge_y=0.2)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  scale_color_brewer(palette="Set2")+
  theme_bw()
```

@@

To verify the analyses are generating expected results, compare the corrected p-values for synonymous mutations to non-synonymous mutations. Unlike non-synonymous mutations, synonymous mutations should not be experiencing selective pressures. Therefore, while there might be differences in the direction and degree of selection on non-synonymous (i.e., p-values close to 0), there should be insignificant differences between clades for synonymous mutations (i.e., p-values close to 1).

```{r verify_analyses}
# Plot the corrected p-values for synonymous and non-synonymous mutations and compare to a significance value of 0.05.
verify_analyses <- comparison_p_values
verify_analyses$synonymous <- ifelse(verify_analyses$c1_synonymous=="TRUE" & verify_analyses$c2_synonymous=="TRUE", "TRUE", "FALSE")
verify_analyses$p_one <- ifelse(verify_analyses$p_value_corrected==1, "P=1", "P<1")

p <- ggplot(verify_analyses, aes(x=p_value_corrected, fill=synonymous, label=..count..))+
  geom_histogram()+
  geom_text(size=2, vjust=-0.5, stat="bin", position="stack")+
  geom_vline(data=data.frame(xint=0.05, p_one="P<1"), aes(xintercept=xint), linetype=2)+
  scale_x_continuous(breaks=c(0,0.25,0.5,0.75,1))+
  labs(x="FDR corrected p-values")+
  theme_bw()

p + ggforce::facet_row(vars(p_one), scales="free", space="free")
```
